A Mechanised Library for Generic (Co)Programming

### 2024-03-07 paper/paper.tex
Some progress on Recursive anamorphisms

\subsection{Recursive Anamorphisms}
Anamorphisms specialise to an hylomorphism into an inductive data type.
Intuitively, this is because a recursive coalgebra can be applied only finitely
many times, so when this is applied to an anamorphism the only possible traces
it can produce are the finite ones.  We show then that an anamorphisms applied
to a recursive coalgebra is in one-to-one correspondence with an hylomorphism
applied to the same recursive coalgebra and the \coq{In} map which is a map into
the inductive data type defined by \coq{In}.

To do this we first define a predicate which holds if and only if the element of
a coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Catamorphisms from this initial algebra are denoted by \coq{ccata}, and they can
be used to define the isomorphism between \coq{{x : GFix F | FinF x}}
and \coq{LFix F}. In particular:
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. We refer the reader to
the supplementary material and Coq code to see how we encode this. In this
paper, however, we focus on the fact that, if anamorphisms produce finite trees,
then their return type can also be the \emph{least} fixed point of the container
extension. We encode in Coq the notion of \emph{recursive anamorphisms} and
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.

@@

\mpav{Moved here}

Note that such proofs will be erased when extracting code to OCaml, and this is
standard practice when building general recursive functions in
Coq\footnote{\dcas{This technique had a name in the community, but I cannot
remember. Something predicates}}. The core definitions are the following:
\begin{coqcode}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

Definition rana_f__ `{setoid A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{coqcode}
Here, \coq{RecF_inv} provides the structurally smaller element for
the Coq recursive definition. We then couple coalgebras with a proof that they
are recursive, and add a coercion from recursive coalgebras to coalgebras.
\begin{coqcode}
Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
\end{coqcode}
In some cases, proving that coalgebras can only be applied finitely many
times is straightforward (e.g.\ they are recursive on some inductive type in
Coq). However, in many cases, proving that a coalgebra is recursive directly
can be tedious and cumbersome. To ease development, we prove that if a
given coalgebra respects a \emph{well-founded relation}, then it is
necessarily recursive.
\begin{coqcode}
Definition respects_relation `{setoid A} (c : Coalg F A)
      {B} (m : A -> B) (R : B -> B -> Prop)
  := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).

Lemma wf_coalg_rec `{setoid A} {B}
  (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
  (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
\end{coqcode}
In the above definition, \coq{m} is used as a form of ``metric''
for termination, and the well-founded relation is on \coq{B}. This
works equivalently to transporting a well-founded relation
\coq{R}
on
\coq{B} to \coq{A} as
\coq{R' := fun x y => R (m x) (m y)}, which is also trivially
well-founded. Using these definitions, we can construct recursive coalgebras by
proving that they respect a well-founded relation, which generally leads to
simpler proofs.

Similarly to the rest of the definitions, we prove that recursive anamorphisms
are respectful, and use this fact to build them as higher order proper
morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
the unique solution to the equation
\coq{f =e l_in \o fmap f \o h},
where \coq{h} is a recursive coalgebra.
\begin{coqcode}
Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...

@@

\section{Extraction}

This section covers how to use our approach to calculate recursive programs
from simple, non-recursive specifications. The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.

>>>>

\subsection{Recursive Anamorphisms}
Anamorphisms specialise to an hylomorphism into an inductive data type.
Intuitively, this is because a recursive coalgebra can be applied only finitely
many times, so when this is applied to an anamorphism the only possible traces
it can produce are the finite ones.  In particular, we can show that
anamorphisms of type \coq{X -> GFix F} applied to a recursive \coq{F}-coalgebra
on \coq{X} are in one-to-one correspondence with hylomorphisms of
type \coq{X -> LFix F} on the same recursive \coq{F}-coalgebra and the
algebra \coq{In} for the inductive data type \coq{LFix F}.

To do this we first define a predicate on elements of coinductive data
structures which holds if and only if the element is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
We can now define catamorphisms denoted by \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
In particular, we can take a catamorphism from the finitary elements
of \coq{GFix F} to the inductive data type \coq{LFix F} using
the \coq{F}-algebra \coq{l_in} and viceversa using catamorphism on the
restriction of \coq{g_in} to the finite elements of \coq{GFix}, which we denote
by \coq{lg_in}.
\begin{coqcode}
Lemma cata_ccata `{setoid A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{setoid A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{coqcode}
Moreover, using this new initial algebra, we can compose catamorphisms and
recursive anamorphisms.

@@

\begin{coqcode}
Definition rana `{eA : setoid A} : RCoalg A ~> A ~> LFix F :=...

@@

\begin{coqcode}
Definition rana_f__ `{setoid A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{coqcode}

% \mpav{Moved here}

% Note that such proofs will be erased when extracting code to OCaml, and this is
% standard practice when building general recursive functions in
% Coq\footnote{\dcas{This technique had a name in the community, but I cannot
% remember. Something predicates}}. The core definitions are the following:
% \begin{coqcode}
% Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

% \end{coqcode}
% Here, \coq{RecF_inv} provides the structurally smaller element for
% the Coq recursive definition. We then couple coalgebras with a proof that they
% are recursive, and add a coercion from recursive coalgebras to coalgebras.
% \begin{coqcode}
% Definition RCoalg F A = Rec {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
% \end{coqcode}
% In some cases, proving that coalgebras can only be applied finitely many
% times is straightforward (e.g.\ they are recursive on some inductive type in
% Coq). However, in many cases, proving that a coalgebra is recursive directly
% can be tedious and cumbersome. To ease development, we prove that if a
% given coalgebra respects a \emph{well-founded relation}, then it is
% necessarily recursive.
% \begin{coqcode}
% Definition respects_relation `{setoid A} (c : Coalg F A)
%       {B} (m : A -> B) (R : B -> B -> Prop)
%   := forall x (p : Pos (shape (c x))), R (m (cont (c x) p)) (m x).

% Lemma wf_coalg_rec `{setoid A} {B}
%   (m : A -> B) (R : B -> B -> Prop) (WF : well_founded R)
%   (c : Coalg F A) (RR : respects_relation c m R) : RecP c.
% \end{coqcode}
% In the above definition, \coq{m} is used as a form of ``metric''
% for termination, and the well-founded relation is on \coq{B}. This
% works equivalently to transporting a well-founded relation
% \coq{R}
% on
% \coq{B} to \coq{A} as
% \coq{R' := fun x y => R (m x) (m y)}, which is also trivially
% well-founded. Using these definitions, we can construct recursive coalgebras by
% proving that they respect a well-founded relation, which generally leads to
% simpler proofs.

% Similarly to the rest of the definitions, we prove that recursive anamorphisms
% are respectful, and use this fact to build them as higher order proper
% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
% the unique solution to the equation
% \coq{f =e l_in \o fmap f \o h},
% where \coq{h} is a recursive coalgebra.
\section{Extraction}
This section covers how to use our approach to calculate recursive programs
from simple, non-recursive specifications. The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.
